Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-17 manuscript/manuscript.tex
section 3

Let $\mathbf{y}_{i}$ be the $J \times 1$ observation vector for subject $i$ such that $y_{ij}$ is the observation for subject $i$ at timepoint $j$. We assume for now that $\mathbf{y}_{i}$ is fully observed. Later, we relax this assumption by allowing for missingness in the components of $\mathbf{y}_{i}$. The general formulation of the multivariate skew normal mixture model is as follows.

$$f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ik} f(\mathbf{y}_i|\boldsymbol\theta_k)$$

@@

In each development class, we assume that a subject's $J\times 1$ outcome vector follows a multivariate skew normal (MSN) distribution. We define a latent variable $Z_i \in \{1,...,K\}$ to denote the cluster membership of subject $i$ for $i = 1,...,n$. We model $\mathbf{y}_{i}|Z_i$ as

$$\mathbf{y}_{i}|Z_i = k \sim MSN_J(\boldsymbol\zeta_k,\boldsymbol\alpha_k,\boldsymbol\Omega_k)$$

@@

We model the effect of covariates on longitudinal development outcomes through the use of a MSN regression model. We adopt a standard stochastic representation of a MSN random variable as the sum of a latent truncated normal random effect and a correlated multivariate normal random error term. The MSN stochastic representation is as follows.
$$\mathbf{y}_{i} = \boldsymbol\beta^T \mathbf{x}_{i} + t_{i}\boldsymbol\psi + \boldsymbol\epsilon_{i}$$

where $\mathbf{x}_i$ is the $P \times 1$ vector of covariate values for subject $i$, $\boldsymbol\beta$ is the $P \times J$ vector of fixed effects coefficients , $t_i \stackrel{iid}{\sim} N_{[0,\infty)}(0,1)$ is a conditional truncated normal random effect, $\boldsymbol\psi$ is the $J \times 1$ vector containing cluster-specific skewness parameters, and $\boldsymbol\epsilon_i \sim N_J(0,\boldsymbol\Sigma_{J \times J})$ is the correlated error term. In can be shown that under these conditions $\mathbf{y}_i$ is multivariate skew normal with location. For conciseness, we present the MSN distribution here outside of the finite mixture context. When using this stochastic representation in our finite mixture model, the parameters $\boldsymbol\beta$, $\boldsymbol\psi$, and $\boldsymbol\Sigma$ are replaced with class-specific counterparts $\boldsymbol\beta_k$, $\boldsymbol\psi_k$, and $\boldsymbol\Sigma_k$.

@@

We note that the multivariate skew normal regression problem can be expressed more succinctly through the use of Matrix Skew Normal distributions. In particular, we define $\mathbf{Y}_{n_k \times J}$ as the response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n(Z_i = k)$ is the total number of observations belonging to class $k$ ($k = 1,...,K$). 
$$\mathbf{Y}_{n_k \times J} \sim \mathcal{MATSN}_{n_k \times J}(\mathbf{M},\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Sigma_k)$$
$$\text{vec}(\mathbf{Y}) = \mathbf{y}_{n_kJ \times 1} = (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})$$
$$\text{vec}(\mathbf{M}) = \boldsymbol\mu_{n_k J \times 1} = (\boldsymbol\mu_1^T,...,\boldsymbol\mu_{n_k}^T)$$

where $\boldsymbol\mu_i = \mathbf{x}_i \boldsymbol\beta_k$ and $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha{kJ})$. This implies that the $i^{th}$ row of $\mathbf{Y}$ is $\mathbf{y}_i \sim \mathcal{MSN}_J(\boldsymbol\mu_i, \boldsymbol\alpha_k, \boldsymbol\Sigma_{J \times J}) = \mathcal{MSN}_J(\mathbf{x}_i \boldsymbol\beta_k, \boldsymbol\alpha_k, \boldsymbol\Sigma)$. Using the stochastic representation presented in Section 3.2, the conditional distribution $\mathbf{y}_i | Z_i, t_i \sim N_J (\mathbf{x}_i \boldsymbol\beta_k + \boldsymbol\psi_k t_i,\boldsymbol\Sigma_k)$. This implies that all responses for cluster $k$ conditioned on $\mathbf{T}_{n_k \times 1}$ is distributed as 
$$\mathbf{Y}_{n_k \times J} | \mathbf{T}_k = \mathbf{t}_k \sim N_{n_k \times J}(\mathbf{M}_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k)$$
$$\text{vec}(\mathbf{M}_k)_{n_kJ \times 1} = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$$\

@@

where $\mathbf{w}_i$ is the vector of class probability covariates for subject $i$, $\boldsymbol\delta_k$ contains the multinomial regression parameters for class $k$, and $K$ is the number of putative clusters specified \textit{a priori}. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under this model, $Z_i|\mathbf{w}_i \sim Multinom.(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{i1},...,\pi_{1K})$. During MCMC estimation, the class labels $z_i$ are updated from their multinomial full conditional distribution and used in the remaining MCMC steps as class assignments.

@@

We allow for missingness of outcomes in the MSN mixture model by imputing missing values from their conditional multivariate normal distributions. We note that
$$\mathbf{y}_i|Z_i,\mathbf{x}_i,t_i,\boldsymbol\beta_{Z_i},\boldsymbol\psi_{Z_i},\boldsymbol\Sigma_{Z_i} \sim N_J(\boldsymbol\beta_{Z_i}^T\mathbf{x}_i  + t_i \boldsymbol\psi_{Z_i}, \boldsymbol\Sigma_{Z_i})$$

This allows us to appeal to standard conditional forms of the multivariate normal distribution to specify the distribution of missing observations. For subject $i$, we use $q_i$ to denote the number of the total $J$ possible repeated measurements are missing. We partition the full $J \times 1$ outcome vector $\mathbf{y}_i$ into the $(J - q_i) \times 1$ observed data vector $\mathbf{y}_i^{obs}$ and the $q_i \times 1$ missing data vector $\mathbf{y}_i^{miss}$. $Y_i = [Y^{miss}_{i_{q \times 1}} | Y^{obs}_{i_{J - q \times 1}}]^T$. We have
$$\mathbf{y}_i^{miss}|\mathbf{y}_i^{obs},\mathbf{x}_i,t_i,Z_i,\boldsymbol\beta_{Z_i},\boldsymbol\psi_{Z_i},\boldsymbol\Sigma_{Z_i} \sim N(\boldsymbol\mu^{miss},\boldsymbol\Sigma^{miss})$$

>>>>

Let $\mathbf{y}_{i}$ be the $J \times 1$ observation vector for subject $i$ such that $y_{ij}$ is the observation for subject $i$ at timepoint $j$. We assume for now that $\mathbf{y}_{i}$ is fully observed. Later, we relax this assumption by allowing for missingness in the components of $\mathbf{y}_{i}$. The general formulation of the multivariate skew normal mixture model is as follows.
\begin{equation}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ik} f(\mathbf{y}_i|\boldsymbol\theta_k)
\end{equation}

@@

In each development class, we assume that a subject's $J\times 1$ outcome vector follows a multivariate skew normal (MSN) distribution. We define a latent variable $Z_i \in \{1,...,K\}$ to denote the cluster membership of subject $i$ for $i = 1,...,n$. We model $\mathbf{y}_{i}|Z_i$ as 
\begin{equation}
\mathbf{y}_{i}|Z_i = k \sim MSN_J(\boldsymbol\zeta_k,\boldsymbol\alpha_k,\boldsymbol\Omega_k)
\end{equation}

@@

We model the effect of covariates on longitudinal development outcomes through the use of a MSN regression model. We adopt a standard stochastic representation of a MSN random variable as the sum of a latent truncated normal random effect and a correlated multivariate normal random error term. The MSN stochastic representation is as follows.
\begin{equation}
\mathbf{y}_{i} = \boldsymbol\beta^T \mathbf{x}_{i} + t_{i}\boldsymbol\psi + \boldsymbol\epsilon_{i}
\end{equation}

where $\mathbf{x}_i$ is the $P \times 1$ vector of covariate values for subject $i$, $\boldsymbol\beta$ is the $P \times J$ vector of fixed effects coefficients , $t_i \stackrel{iid}{\sim} N_{[0,\infty)}(0,1)$ is a conditional truncated normal random effect, $\boldsymbol\psi$ is the $J \times 1$ vector containing cluster-specific skewness parameters, and $\boldsymbol\epsilon_i \sim N_J(0,\boldsymbol\Sigma)$ is the correlated error term. In can be shown that under these conditions $\mathbf{y}_i$ is multivariate skew normal with location. For conciseness, we present the MSN distribution here outside of the finite mixture context. When using this stochastic representation in our finite mixture model, the parameters $\boldsymbol\beta$, $\boldsymbol\psi$, and $\boldsymbol\Sigma$ are replaced with class-specific counterparts $\boldsymbol\beta_k$, $\boldsymbol\psi_k$, and $\boldsymbol\Sigma_k$.

@@

We note that the multivariate skew normal regression problem can be expressed more succinctly through the use of Matrix Skew Normal distributions. In particular, we define $\mathbf{Y}_{n_k \times J}$ as the response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n(Z_i = k)$ is the total number of observations belonging to class $k$ ($k = 1,...,K$). 
$$\mathbf{Y}_{n_k \times J} \sim MatSN_{n_k \times J}(\mathbf{M},\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Sigma_k)$$
$$\text{vec}(\mathbf{Y}) = \mathbf{y}_{n_kJ \times 1} = (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})$$
$$\text{vec}(\mathbf{M}) = \boldsymbol\mu_{n_k J \times 1} = (\boldsymbol\mu_1^T,...,\boldsymbol\mu_{n_k}^T)$$

where $\boldsymbol\mu_i = \mathbf{x}_i \boldsymbol\beta_k$ and $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})$. This implies that the $i^{th}$ row of $\mathbf{Y}$ is $\mathbf{y}_i \sim MSN_J(\boldsymbol\mu_i, \boldsymbol\alpha_k, \boldsymbol\Sigma_k) = {MSN}_J(\mathbf{x}_i \boldsymbol\beta_k, \boldsymbol\alpha_k, \boldsymbol\Sigma_k)$. Using the stochastic representation presented in Section 3.2, the conditional distribution of $\mathbf{y}_i | Z_i, t_i$ is $N_J (\mathbf{x}_i \boldsymbol\beta_k + \boldsymbol\psi_k t_i,\boldsymbol\Sigma_k)$. This implies that all responses for cluster $k$ conditioned on $\mathbf{T}_{n_k \times 1}$ is distributed as 
$$\mathbf{Y}_{n_k \times J} | \mathbf{T}_k = \mathbf{t}_k \sim N_{n_k \times J}(\mathbf{M}_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k)$$
$$\text{vec}(\mathbf{M}_k)_{n_kJ \times 1} = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$$\

@@

where $\mathbf{w}_i$ is the vector of class probability covariates for subject $i$, $\boldsymbol\delta_k$ contains the multinomial regression parameters for class $k$, and $K$ is the number of putative clusters specified \textit{a priori}. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under this model, $Z_i|\mathbf{w}_i \sim Multinom(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{i1},...,\pi_{1K})$. During MCMC estimation, the class labels $z_i$ are updated from their multinomial full conditional distribution and used in the remaining MCMC steps as class assignments.

@@

We allow for missingness of outcomes in the MSN mixture model by imputing missing values from their conditional multivariate normal distributions. We note that
$$\mathbf{y}_i|Z_i,\mathbf{x}_i,t_i,\boldsymbol\beta_{k},\boldsymbol\psi_{k},\boldsymbol\Sigma_{k} \sim N_J(\boldsymbol\beta_{k}^T\mathbf{x}_i  + t_i \boldsymbol\psi_{k}, \boldsymbol\Sigma_{k})$$

where $Z_i = k$. This allows us to appeal to standard conditional forms of the multivariate normal distribution to specify the distribution of missing observations. For subject $i$, we use $q_i$ to denote the number of the total $J$ possible repeated measurements are missing. We partition the full $J \times 1$ outcome vector $\mathbf{y}_i$ into the $(J - q_i) \times 1$ observed data vector $\mathbf{y}_i^{obs}$ and the $q_i \times 1$ missing data vector $\mathbf{y}_i^{miss}$. $Y_i = [Y^{miss}_{i_{q \times 1}} | Y^{obs}_{i_{J - q \times 1}}]^T$. We have
$$\mathbf{y}_i^{miss}|\mathbf{y}_i^{obs},\mathbf{x}_i,t_i,Z_i,\boldsymbol\beta_{k},\boldsymbol\psi_{k},\boldsymbol\Sigma_{k} \sim N(\boldsymbol\mu^{miss},\boldsymbol\Sigma^{miss})$$
